Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-07-12 Manuscript/manuscript.Rmd
intro edits

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking]. Studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological].  However, the growing number of publically available annotated genomic elements and there collinearity pose a significant problem for establishing methods robust enough to determine the most important epigenomic factors that contribute to TAD boudandary formation.

@@

First, the use of different levels of association between TAD boundary regions and genomic features as predictors remain underexplored. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association as it simply counts the total number of peaks defining a genomic annotation that are inside a given TAD boundary region. Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of overlap betwee a genomic annotation and a boundary region. However, the distance between boundary regions and genomic annotations are currently unexplored. This association allows for quantifying the spatial relationship between genomic elements and boundary regions. We hypothesize that the inclusion of this type of association will correspond with more accurate identification of functional genomic elements association with TAD boundary formation.

Second, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These boundary regions make up the minority class, while the non-boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as twice the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. That is, the PI is bounded by 0 and 1. A dataset is considered balanced if PI$\approx$1, and imbalanced if PI$<<$1. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. We hypothesize that the implementation of novel data resampling techniques will significantly improve model performance. Futhermore, the choice of performance metric can mask the poor performance of models implemented on imbalanced data.  It has been shown that conventional metrics such as AUC and accuracy are insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. Therefore, we hypothesize that a more composite metric such as Matthews Correlation Coefficient (MCC), might provide more sensitive assessment of model performance, given that it penalizes for both false positives and false negatives.

@@

Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, was developed to predict TAD boundary regions using a collection of 69 transcription factor binding sites on the GM12878 cell line [@mourad2016computational].

@@

In this study, we developed a computational framework aimed at determining the optimal set of data level characteristics necessary for optimal prediction of TAD boundaries and identification of influential genomic annotations. We utilize random forest classification models built on HiC-data from two cell lines, including GM12878 and K562, obtained at four different resolutions, including 10 kb, 25 kb, 50 kb, and 100 kb. For each cell line and resolution, we compared three feature engineering techniques, including counts, percents, and distances. Additionally, we compared three re-sampling techniques, including random under-sampling, random over-sampling, and a weighted combination of both referred to as Synthetic Minority Over-sampling Technique (SMOTE). Model performances were evaluated and compared using Matthews Correlation Coefficient (MCC).

[???Include concluding paragraph]

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking]. Studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological]. However, the growing number of publically available annotated genomic elements and there collinearity pose a significant problem for establishing methods robust enough to determine the most important epigenomic factors that contribute to TAD boudandary formation. Moreover, the plasticity of functional annotations make TAD boundary prediction a cell line specific problem [@wang2012widespread]

@@

First, the use of different levels of association between TAD boundary regions and genomic features as predictors remain underexplored. These associations are termed predictor-types and make up the feature space of classifiers. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association as it simply counts the total number of peaks defining a genomic annotation that are inside a given TAD boundary region. Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of overlap betwee a genomic annotation and a boundary region. However, the distance between boundary regions and genomic annotations are currently unexplored. This association allows for quantifying the spatial relationship between genomic elements and boundary regions. We hypothesize that the inclusion of this type of association will correspond with more accurate identification of functional genomic elements association with TAD boundary formation.

Secondly, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These boundary regions make up the minority class, while the non-boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as twice the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. That is, the PI is bounded by 0 and 1. A dataset is considered balanced if PI $\approx$ 1, and imbalanced if PI$<<$1. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. We hypothesize that the implementation of novel data resampling techniques will significantly improve model performance. Futhermore, the choice of performance metric can mask the poor performance of models implemented on imbalanced data.  It has been shown that conventional metrics such as AUC and accuracy are insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. Therefore, we hypothesize that a more composite metric such as Matthews Correlation Coefficient (MCC), might provide more sensitive assessment of model performance, given that it penalizes for both false positives and false negatives.

@@

<!--[still revising]-->
Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, was developed to predict TAD boundary regions using a collection of 69 transcription factor binding sites on the GM12878 cell line [@mourad2016computational].

@@

In this study, we have developed a computational framework aimed at determining the optimal set of data level characteristics necessary for optimal prediction of TAD boundaries and identification of influential genomic annotations. These include the best combination of predictor type, resampling technique, and data resolution. We utilize random forest classification models built on Hi-C data from two cell lines, including GM12878 and K562, obtained at four different resolutions, including 10 kb, 25 kb, 50 kb, and 100 kb. For each cell line and resolution, we compared three predictor types, including counts, percents, and distances. Additionally, we compared three re-sampling techniques, including random under-sampling, random over-sampling, and a weighted combination of both referred to as Synthetic Minority Over-sampling Technique (SMOTE). Model performances were evaluated and compared using Matthews Correlation Coefficient (MCC). We performed 10-fold cross-validation for parameter tuning and to guard against over-fitting. We found that the optimally built cell line specific model is one implemented using distance type predictors, with SMOTE resampling, at the 10-kb data resolution. We also determined that simply including transcription factor binding sites was necessary for optimal performance. The inclusion of chromatin states and histone modifications was found to induce noise into the model. Specifically, we found that the transcription factors CTCF, ZNF143, RAD21, and SMC3 exhibited high predictive importances, confirming there involvement in TAD boundary formation. We then aimed to evaluate the cell line specificity of TAD boundaries by validating our optimally built models across cell lines. We found that models built and validated on the same cell line outperformed models that were validated across cell lines, confirming the cell line specific nature and plasticity of TAD boundaries and functional genomic annotations respectively. <!--[add conclusion from results of predicting finer bins]-->
